---
title: "SNP Association - Asthma"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

This analysis aims to identify SNPs that are associated with individuals that have asthma. Three methods will be used:
- Genetic association using individual SNPs
- Halotype association
- Polygenic risk scores

```{r}
library(SNPassoc)
```

Load the data, originally from https://github.com/isglobal-brge/brgedata/blob/master/data/asthma.rda

```{r}
load("./data/asthma.rda")
head(asthma)
```

## Genetic Association Studies

Take a closer look at the features.

```{r}
str(asthma, list.len=9)
```

Identify which columns contain the SNP data.

```{r}
 asthma.s <- setupSNP(data=asthma, colSNPs=7:ncol(asthma), sep="")
```

Let's take a look at one of those columns.

```{r}
 head(asthma.s$rs1422993)
```
```{r}
 class(asthma.s$rs1422993)
```

Summarise the content of this SNP.

```{r}
summary(asthma.s$rs1422993)
```


```{r}
 head(summary(asthma.s))
```

See where the missing values and if there are any patterns here.

```{r}
 plotMissing(asthma.s, print.labels.SNPs = FALSE)
```

See which SNPs do not meet the Hardy-Weinberg equilibrium.

```{r}
hwe <- tableHWE(asthma.s, casecontrol)

# Divide by 51 as that's the number of SNPs we are analysing.
threshold <- 0.05 / 51
snpNHWE <- hwe[,2] < threshold
rownames(hwe)[snpNHWE]
hwe[snpNHWE,]
```

All SNPs meet the HWE in the control group, so none of them need to be removed from the dataset before further analysis.

For a bit of fun, let's find the SNP that is most outside the HWE.

```{r}
which.min(hwe[,3])
```

Perform an association analysis between rs184448 and the casecontrol.

```{r}
association(casecontrol ~ rs184448, data = asthma.s)
```

The dominant model is the most statistically significant, as individuals with this SNP seem to be 55% more likely to have asthma if they have this mutation.

```{r}
maxstat(asthma.s$casecontrol, asthma.s$rs184448)
```

When using the maximum statistic, the result has a low p-value, but it is still above our thresold (~0.0098).

Control for a few variables to see how that affects the significance.

```{r}
association(casecontrol ~ rs184448 + gender + smoke + bmi, asthma.s)
```

The p-value has decreased and is now close to our threshold.

To further inspect the data, we can take a look at the association when smoking is modeled as a factor.

```{r}
association(casecontrol ~ dominant(rs184448)*factor(smoke), data=asthma.s)
```

Non-smokers with this mutation have a higher association of asthma than non-smokers, although this result is not statistically significant. 

Expand the analysis to perform a whole genome association analysis using all of the SNPs.

```{r}
ans <- WGassociation(casecontrol, data=asthma.s)
head(ans)
```

Plot the results.

```{r}
plot(ans)
```

None of the SNPs meet our significance threshold.

```{r}
ans.max <- maxstat(asthma.s, casecontrol)
ans.max[,ans.max["Pr(>z)",] <= threshold]
```

## Haplotype Analysis

Perform a haplotype analysis to find sets of alleles that are associated with asthma.

```{r}
library(LDheatmap)
library(genetics)
library(biomaRt)
library(haplo.stats)
library(plyr)
```

Connect to the BioMart database to get the annotations for human SNPs.

```{r}
mart <- useMart("ENSEMBL_MART_SNP", dataset = "hsapiens_snp")
nrow(listFilters(mart))
head(listFilters(mart))
```

Idenitfy the row that contains the annotations.

```{r}
listFilters(mart)[11,]
```

Get the relevant attributes from the BioMart database.

```{r}
snps <- labels(asthma.s)
snpInfo <- getBM(
  c("refsnp_id", "chr_name", "chrom_start", "allele"),
  filters=c("snp_filter"),
  values=snps,
  mart=mart)

head(snpInfo)
```

Count the occurrences of each chromosome. 

```{r}
count(snpInfo, "chr_name")
```

Find the min and max loci of the SNPs on chromosome 7.

```{r}
min(snpInfo[snpInfo$chr_name == 7,]$chrom_start)
max(snpInfo[snpInfo$chr_name == 7,]$chrom_start)
```

Filter the snpInfo to the relevant data and set up the haplotype data.

```{r}
mask <- with(snpInfo, chr_name=="7" & chrom_start>34.6e6 & chrom_start<34.9e6)
snps.sel <- snpInfo[mask, "refsnp_id"]
sel <- which(names(asthma)%in%snps.sel)
asthma.hap <- setupSNP(asthma, sel, sep="")
```

Create a genotype object.

```{r}
snp.pos <- snpInfo[mask, "chrom_start"]
snp.geno <- data.frame(lapply(asthma.hap[, snps.sel], genotype))
```

Create an LD heatmap.

```{r}
LDheatmap(
  snp.geno,
  LDmeasure = "r",
  title = "Pairwise LD in r^2",
  add.map = TRUE,
  name = "myLDgrob",
  add.key = TRUE,
  SNP.name=snps.sel,
)
```

Select a set of neighbouring SNPs and compute their haplotype frequency.

```{r}
snpsH <- c("rs10486657", "rs10250709", "rs10238983", "rs6958905")
genoH <- make.geno(asthma.hap, snpsH)
em <- haplo.em(genoH, locus.label = snpsH, miss.val = c(0, NA))
em
```

Haplotype 6 is the most probable, accumulating nearly 50% of the haplotype sample.

Measure the association between any of the haplotypes and an individual having asthma using a GLM.

```{r}
trait <- asthma.hap$casecontrol
mod <- haplo.glm(
  trait ~ genoH,
  family="binomial",
  locus.label=snpsH,
  allele.lev=attributes(genoH)$unique.alleles,
  control = haplo.glm.control(haplo.freq.min=0.05)
)
intervals(mod)
```

None of the haplotypes are significantly associated with asthma.

Rather than selecting four SNPs like above, we can use a sliding window for better haplotype inference. We'll use a window size between 3 and 6.


```{r}
snpsH2 <- colnames(snp.geno)
genoH2 <- make.geno(asthma.hap, snpsH2)
trait <- asthma.hap$casecontrol
haplo.score <- list()

for (i in 3:6) {
  haplo.score[[i-2]] <- haplo.score.slide(
    trait,
    genoH2,
    trait.type="binomial",
    n.slide=i,
    simulate=TRUE,
    sim.control=score.sim.control(min.sim=100, max.sim=200)
  )
}
```

Plot the results.

```{r}
par(mfrow=c(2,2))

for (i in 3:6) {
  plot(haplo.score[[i-2]])
  title(paste("Sliding Window=", i, sep=""))
}
```

The highest -log10 p-value is obtained for a haplotype of 3 SNP length starting at the 9th SNP of the selected SNPs. Let's create another GLM using this haplotype to see if we can find a statistically significant association with asthma.

```{r}
snpsH3 <- snpsH2[9:11]
genoH3 <- make.geno(asthma.hap, snpsH3)
mod <- haplo.glm(
  trait~genoH3,
  family="binomial",
  locus.label=snpsH3,
  allele.lev=attributes(genoH3)$unique.alleles,
  control = haplo.glm.control(haplo.freq.min=0.05)
)

intervals(mod)
```

We see that individuals carrying the haplotype GCA have a 35% increased risk of asthma relative to those having the reference haplotype ACG.

```{r}
lrt <- mod$lrt
pchisq(lrt$lrt, lrt$df, lower=FALSE)
```

The low chi-squared value suggests that this difference is significant.

## Genetric Score

Create a polygenic risk score to assess the collective prediction of the phenotype risk associated with multiple SNPs.

```{r}
library(PredictABEL)
```

Identify which columns in the dataset contains SNPs and measure their association with asthma using a whole genome association analysis. 

```{r}
dd.s <- setupSNP(asthma, 7:ncol(asthma), sep="")
ans <- WGassociation(casecontrol, dd.s, model="log-additive")
```

Filter the SNPs so that only those that show an association at 0.1 level are included in the model.

```{r}
sel <- labels(dd.s)[additive(ans) < 0.1]
dd.sel <- dd.s[,sel]
head(dd.sel)
```

Reformat the SNP genotypes as 0, 1, and 2 using additive so that they can be used as a feature in the model. Join the case/control variables and the reformatted SNP genotypes.

```{r}
dd.sel <- data.frame(lapply(dd.sel, additive))
dd.end <- data.frame(casecontrol=dd.s$casecontrol, dd.sel)
head(dd.end)
```

First remove SNP genotypes with missing values, as they are not supported by the model. Then perform a stepwise model selection to determine which SNPs best predict asthma status.

```{r}
dd.end.complete <- dd.end[complete.cases(dd.end),]
mod <- stepAIC(
  glm(casecontrol ~ ., dd.end.complete, family="binomial"),
  method="forward",
  trace=0
)

snps.score <- names(coef(mod))[-1]
snps.score
```

Collectively, these are the most predictive SNPs. They will be used to create the genetic score. But first, let's take a look at a summary that shows the SNPs’ associations of the selected SNPs with asthma.

```{r}
summary(mod)
```

Identify the columns of the SNPs that will be used to compute the genetic score.

```{r}
pos <- which(names(dd.end.complete)%in%snps.score)
pos
```

Create the scores using riskScore.

```{r}
score <- riskScore(
  mod,
  data=dd.end.complete,
  cGenPreds=pos,
  Type="unweighted"
)
table(score)
```

Plot the distribution of the risk scores.

```{r}
hist(score, col="gray90")
```

Test the association between the genetic score and asthma status using a generalised linear model.

```{r}
mod.lin <- glm(casecontrol~score, dd.end.complete, family="binomial")
exp(coef(mod.lin)[2])
```

According to this model, the risk of asthma increases 21% per risk allele that an individual has.

Plot the model's performance.

```{r}
predrisk <- predRisk(mod.lin, dd.end.complete)
plotROC(data=dd.end.complete, cOutcome=1, predrisk = predrisk)
```

