---
title: "SNP Association - Asthma"
output:
  html_document:
    df_print: paged
---

This analysis aims to identify SNPs that are associated with individuals that have asthma.

```{r}
library(SNPassoc)
```

Load the data, originally from https://github.com/isglobal-brge/brgedata/blob/master/data/asthma.rda

```{r}
load("./data/asthma.rda")
head(asthma)
```
Take a closer look at the features.

```{r}
str(asthma, list.len=9)
```

Identify which columns contain the SNP data.

```{r}
 asthma.s <- setupSNP(data=asthma, colSNPs=7:ncol(asthma), sep="")
```

Let's take a look at one of those columns.

```{r}
 head(asthma.s$rs1422993)
```
```{r}
 class(asthma.s$rs1422993)
```

Summarise the content of this SNP.

```{r}
summary(asthma.s$rs1422993)
```


```{r}
 head(summary(asthma.s))
```

See where the missing values and if there are any patterns here.

```{r}
 plotMissing(asthma.s, print.labels.SNPs = FALSE)
```

See which SNPs do not meet the Hardy-Weinberg equilibrium.

```{r}
hwe <- tableHWE(asthma.s, casecontrol)

# Divide by 51 as that's the number of SNPs we are analysing.
threshold <- 0.05 / 51
snpNHWE <- hwe[,2] < threshold
rownames(hwe)[snpNHWE]
hwe[snpNHWE,]
```

All SNPs meet the HWE in the control group, so none of them need to be removed from the dataset before further analysis.

For a bit of fun, let's find the SNP that is most outside the HWE.

```{r}
which.min(hwe[,3])
```

Perform an association analysis between rs184448 and the casecontrol.

```{r}
association(casecontrol ~ rs184448, data = asthma.s)
```

The dominant model is the most statistically significant, as individuals with this SNP seem to be 55% more likely to have asthma if they have this mutation.

```{r}
maxstat(asthma.s$casecontrol, asthma.s$rs184448)
```

When using the maximum statistic, the result has a low p-value, but it is still above our thresold (~0.0098).

Control for a few variables to see how that affects the significance.

```{r}
association(casecontrol ~ rs184448 + gender + smoke + bmi, asthma.s)
```

The p-value has decreased and is now close to our threshold.

To further inspect the data, we can take a look at the association when smoking is modeled as a factor.

```{r}
association(casecontrol ~ dominant(rs184448)*factor(smoke), data=asthma.s)
```

Non-smokers with this mutation have a higher association of asthma than non-smokers, although this result is not statistically significant. 

Expand the analysis to perform a whole genome association analysis using all of the SNPs.

```{r}
ans <- WGassociation(casecontrol, data=asthma.s)
head(ans)
```

Plot the results.

```{r}
plot(ans)
```

None of the SNPs meet our significance threshold.

```{r}
ans.max[,ans.max["Pr(>z)",] <= threshold]
```

Try to find significance once more after controlling for a few variables.

```{r}
ans2 <- WGassociation(casecontrol ~ gender + smoke + bmi, data=asthma.s)
```


```{r}
plot(ans)
```

Still nothing as none of the p-value cross the significance threshold with the Bonferroni adjustment.


